---
title: 'Final Project'
subtitle: "PSTAT122: Design and Analysis of Experiments" 

author:
  - name: "Winter 2026"
   # affiliations:
    #  - name: "Fall 2025"
affiliation-title: "Quarter"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    geometry: margin=0.2in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)

```

:::: callout
::: {style="text-align: center"}
[ **STUDENT NAME** ]{style="color: blue;"}
:::

-   Zachary Scholnick (3374824)
-   Anjali Dev (A167A78)
-   Logan Marks (9536947)

::::

::: callout-caution
# Due Date

**Due Date:** Monday, March 16, 2026, 11:59 PM
:::

```{r}
library(agricolae) 
library(car)         
library(dplyr)  
library(ggplot2)
library(knitr)
```

# Introduction

In the field of map-making and cartography, scale is one of the most critical and challenging decisions practitioners face. At too large of a scale (zoomed in), key features are easily visible, but lose out on important context. At the same time, a very low scale (zoomed out), means that the information is difficult to see and fine spatial details are lost. Getting the scale of your map right is critical in conveying key information - and for that reason it's a question asked every time a map is created. On top of the practical aspects of mapmaking, humans possess a number of cognitive biases that make interpretations of proportion and area very difficult. As such, this experiment was motivated by the desire to provide real insights into what scale of representation makes estimation of area most intuitive.

While tackling this in entirety is beyond the scope of this project, we chose to focus on testing a key component of this issue: How does the scale of a mapped polygon affect the accuracy in which users are able to estimate the area? Do people systematically overestimate or underestimate polygon area at different scales?


# Experimental Design

To assess this, we created 3 polygons of equal area, 1.71 square miles, selected because mapped parcels, agricultural units, and other targets of interest are commonly in the 1-3 square mile range. We also opted to map polygons of equal area to avoid introducing an additional confounding variable. These 3 polygons represented common archetypes in geometric representation: the simple polygon, the rounded polygon, and the more complex, concave polygon. We selected 3 random regions in the Patagonian Steppe in Argentina, as this region lacked easily recognizable reference points of scale, like trees or streams, and consisted mainly of rock formations of unidentifiable size. To ensure consistent areal calculation of mapped region and polygon, we opted for the POSGAR 2007/EPSG:5343 projection, which is the official geodetic reference system of Argentina.

Each polygon was then plotted at the scale ratios of 1:13,390, 1:26,780, and 1:59,852 to provide a large-scale, medium-scale, and small-scale plot. In a more generalizable interpretation, this means that we plotted polygons such that they covered 40%, 10%, and 2% of the entire mapped region. The scale bar itself was dynamically recalculated using QGIS based on the current map extent with a fixed display size for a constant visual across maps.

| Scale Label | Scale Ratio | Polygon Coverage |
|-------------|-------------|------------------|
| Large (L)   | 1:13,390    | 40%              |
| Medium (M)  | 1:26,780    | 10%              |
| Small (S)   | 1:59,852    | 2%               |

: Scale treatments and polygon coverage of map frame {#tbl-scales}

Since our project was trying to capture natural, intuitive estimates of area, participants were each shown a unique combination of scale and polygon type to prevent prior exposure from influencing estimates. Each participant receieved 3 treatment assignments, randomized using a Latin square structure to ensure that each participant only saw each scale level and polygon type exactly once in varying orders.

We expected that significant variation in spatial cognitive ability across tested individuals would be a major driver in estimation accuracy. By modeling participant as a random effect, we capture the inter-participant variability and generalizes our insights to a wider population. We also hypothesized that polygon shape would alter estimation difficulty regardless of scale. By introducing 3 archetypal polygons as a fixed effect, we can control for this variability to better isolate the effect of scale.

::: {style="display: flex; justify-content: center"}
|                   | Curved Polygon | Simple Polygon | Concave Polygon |
|-------------------|----------------|----------------|-----------------|
| **Participant 7** | S              | M              | L               |
| **Participant 12**| M              | L              | S               |
| **Participant 2** | L              | S              | M               |

: Example Latin square treatment assignment for one replicate block {#tbl-latin}
:::

In this experiment, each participant was given 30 seconds per map to write down an area estimate in square miles of a curved, simple and concave polygon, each shown at a unique scale. This response was then transformed into a signed percent error:

$$\text{Percent Error} = \frac{\hat{a} - a}{a} \times 100$$

where $\hat{a}$ is the participant's estimate and $a = 1.71$ square miles is the true polygon area.

To formalize our experiment into a statistical model for analysis, we implemented a two-way ANOVA with blocking on participant. The response variable $Y_{ijk}$ denotes the signed percent error for participant $k$ receiving scale treatment $i$ and polygon type $j$. The model is:

$$Y_{ijk} = \mu + \tau_i + \beta_j + \rho_k + \epsilon_{ijk}$$

- $i = 1, 2, 3$ indexes scale treatment (Large, Medium, Small)
- $j = 1, 2, 3$ indexes polygon type (simple, curved, concave)
- $k = 1, \cdots, 15$ indexes participant block
- $\mu$ is the overall mean percent error
- $\tau_i$ is the fixed effect of scale treatment $i$, with $\sum \tau_i = 0$
- $\beta_j$ is the fixed effect of polygon type $j$, with $\sum \beta_j = 0$
- $\rho_k$ is the fixed effect of participant block $k$, with $\sum \rho_k = 0$
- $\epsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)$ is the residual error



```{r}
#### Assigning Randomized + Balanced Treatments for Participants ####
library(agricolae)

scales <- c("L", "M", "S")
polygon_types <- c("curved", "simple", "concave")

# randomly assigning each participant to a specific replicate + row
set.seed(42)
participant_assignment <- data.frame(
  replicate = rep(1:5, each = 3),
  row = rep(1:3, times = 5),
  # Randomly assigning participants through these slots
  participant = paste0("P", sample(1:15)) 
)

# Generating the 5 Latin Squares
design_df <- do.call(rbind, lapply(1:5, function(i) {
  book <- design.lsd(scales, seed = i * 10)$book
  book$replicate <- i
  
  # Assign polygon types to columns (1, 2, 3)
  set.seed(i)
  col_map <- setNames(sample(polygon_types), 1:3)
  book$polygon <- col_map[as.character(book$col)]
  
  return(book)
}))

# Merge on BOTH replicate and row to build final design dataframe
design_df <- merge(design_df, participant_assignment, by = c("replicate", "row"))

# 4. Sort neatly by participant (P1 to P15) and then by trial/column (1, 2, 3) to make implementing easy
design_df <- design_df[order(as.numeric(gsub("P", "", design_df$participant)), design_df$col), 
                       c("participant", "replicate", "row", "col", "scales", "polygon")]

# Creating an empty response column
design_df$percent_error <- NA

cat("\nFull Design\n")
print(design_df)
```

# Data Collection

**Procedure** 
Each participant was given 30 seconds to estimate the area in square miles of each polygon shown. Trials were conducted using a slide presentation displayed on a laptop screen, with each slide showing one polygon at its assigned scale alongside a scale bar. Maps were shown on the same monitor at the same display size to keep stimuli constant. Participants recorded their estimates on paper. All 5 replicates were conducted in the UCSB library on March 14, 2026. Participants were seated individually and shown the slides in sequence according to the generated Latin Squares, with no prior knowledge of the true polygon area or that all three polygons were equal in size.

ADD AS APPENDIX Slides for running trials: https://docs.google.com/presentation/d/1cQkHKFSfIFLclCvy6DIQ8lRVRss23OmEZ6OrBmpAJ2I/edit?usp=sharing

**Challenges/Adjustments**
 Several challenges arose during data collection. The most notable was that some participants were unfamiliar with square miles as a unit of measurement, which may have introduced systematic bias into estimates regardless of scale treatment. While participants were not coached on how to use the scale bar, some may have relied on intuition rather than calculation. Additionally, the 30-second time limit, while intentional in capturing instinctive responses, may have prevented participants from making more deliberate estimates — meaning results may reflect first impressions of area rather than considered judgments. 

Finally, as trials were conducted in a shared library space, there is a very small possibility that some participants overheard the estimates of others, which could have anchored subsequent guesses. Protocol was updated so that participants could not use outside props like a pencil, phone, or keys to match distances from the scale bar onto the map. This was to ensure constant conditions in estimating area across trials.

**Data Presentation**

```{r}
enter_data <- function(df, p_name, estimates) {

  # calculate the percent error based on the true area (1.71 sq miles)
  errors <- ((estimates - 1.71) / 1.71) * 100
  
  for(i in 1:3) {
    df$percent_error[df$participant == p_name & df$col == i] <- errors[i]
  }
  
  return(df)
}

#data collection 
design_df <- enter_data(design_df, "P1", c(4.8, 4.5, 3.0))
design_df <- enter_data(design_df, "P2", c(3, 3, 3))
design_df <- enter_data(design_df, "P3", c(2, 1, 3.2))
design_df <- enter_data(design_df, "P4", c(3, 4, 4.5))
design_df <- enter_data(design_df, "P5", c(3.1, 2.8, 2.6))
design_df <- enter_data(design_df, "P6", c(2.4, 2.2, 1.6))
design_df <- enter_data(design_df, "P7", c(5, 12, 6))
design_df <- enter_data(design_df, "P8", c(2.4, 2.5, 2))
design_df <- enter_data(design_df, "P9", c(1.43, 2.8, 3.2))
design_df <- enter_data(design_df, "P10", c(3.2, 3.5, 4.1))
design_df <- enter_data(design_df, "P11", c(2.7, 2.9, 3.1))
design_df <- enter_data(design_df, "P12", c(4.3, 4.1, 3.9))
design_df <- enter_data(design_df, "P13", c(3.6, 3.2, 3.8))
design_df <- enter_data(design_df, "P14", c(2.9, 2.4, 2.7))
design_df <- enter_data(design_df, "P15", c(4.8, 4.5, 4.9))

knitr::kable(design_df, 
             caption = "Full experimental data with percent error by participant, polygon type, and scale",
             digits = 2)
```


# Analysis

**Exploratory Data**:
```{r}
library(dplyr)
library(ggplot2)

# Basic summary statistics by scale
summary_stats <- design_df %>%
  group_by(scales) %>%
  summarise(
    Mean = round(mean(percent_error, na.rm = TRUE), 2),
    SD = round(sd(percent_error, na.rm = TRUE), 2),
    Min = round(min(percent_error, na.rm = TRUE), 2),
    Max = round(max(percent_error, na.rm = TRUE), 2),
    N = n()
  )

knitr::kable(summary_stats, caption = "Summary statistics of percent error by scale treatment")
```


```{r}
# Boxplot by Scale Treatment
ggplot(design_df, aes(x = scales, y = percent_error, fill = scales)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_cartesian(ylim = c(-100, 300)) +
  labs(title = "Percent Error by Scale Treatment",
       x = "Scale", y = "Percent Error (%)") +
  theme_minimal()
```

Note: 1 observation of 601% error on the medium scale was omitted for visual clarity
```{r}
# Boxplot by polygon type
ggplot(design_df, aes(x = polygon, y = percent_error, fill = polygon)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_cartesian(ylim = c(-100, 300)) +
  labs(title = "Percent Error by Polygon Type",
       x = "Polygon Type", y = "Percent Error (%)") +
  theme_minimal()
```

Note: 1 observation of 601% error on the curved polygon was omitted for visual clarity

The boxplots show that percent error is fairly similar across all three scale treatments (Large, Medium, Small), with most participants overestimating the polygon area by around 60–100%. There is one clear outlier with a percent error near 600%, which heavily influences the spread. Across polygon types, concave polygons had the most consistent estimates while curved polygons showed the most variation. Overall, the differences between groups appear modest, which we will be tested formally.

**Hypothesis Testing**: Test your hypothesis with an appropriate statistical test (e.g., ANOVA).

We test the following hypotheses for scale treatment:
$$H_0: \tau_L = \tau_M = \tau_S = 0$$
$$H_a: \tau_i \neq 0 \text{ for at least one } i$$

and for polygon type:
$$H_0: \beta_{\text{simple}} = \beta_{\text{curved}} = \beta_{\text{concave}} = 0$$
$$H_a: \beta_j \neq 0 \text{ for at least one } j$$

where $\tau_i$ and $\beta_j$ are the fixed effects of scale and polygon type as defined in the model above, and $Y_{ijk}$ is the signed percent error response. In context, $H_0$ for scale proposes that the geographic scale at which a polygon is displayed has no effect on the accuracy of area estimation. Equivalently, it proposes that participants estimate equally well regardless of whether the polygon occupies 40\%, 10\%, or 2\% of the map frame. $H_0$ for polygon type asserts that polygon shape has no independent effect on estimation accuracy after controlling for scale and participant. While the participant blocking factor $\rho_k$ is included to contain the individual variations in spatial reasoning ability, it's not the primary interest.

**Tables, Figures, & Code**:
```{r echo=FALSE}
# fit models and run diagnostics

# percent error — full data
mod_pct <- aov(percent_error ~ scales + polygon + participant, data = design_df)

sw_pct       <- shapiro.test(residuals(mod_pct))
design_df$resid_pct <- residuals(mod_pct)
lev_scale_pct   <- leveneTest(resid_pct ~ scales, data = design_df)
lev_poly_pct    <- leveneTest(resid_pct ~ polygon, data = design_df)
design_df$std_resid <- residuals(mod_pct) / sigma(mod_pct)
outliers <- design_df %>%
  filter(abs(std_resid) > 2.5) %>%
  select(participant, scales, polygon, percent_error, std_resid)

# percent error — P7 removed
df_noP7 <- design_df %>% filter(participant != "P7")
mod_noP7 <- aov(percent_error ~ scales + polygon + participant, data = df_noP7)

sw_noP7      <- shapiro.test(residuals(mod_noP7))
df_noP7$resid_noP7 <- residuals(mod_noP7)
lev_scale_noP7  <- leveneTest(resid_noP7 ~ scales, data = df_noP7)
lev_poly_noP7   <- leveneTest(resid_noP7 ~ polygon, data = df_noP7)

# log ratio — full data
design_df$log_ratio <- log(design_df$percent_error / 100 + 1)
mod_log <- aov(log_ratio ~ scales + polygon + participant, data = design_df)

sw_log       <- shapiro.test(residuals(mod_log))
design_df$resid_log <- residuals(mod_log)
lev_scale_log   <- leveneTest(resid_log ~ scales, data = design_df)
lev_poly_log    <- leveneTest(resid_log ~ polygon, data = design_df)
design_df$std_resid_log <- residuals(mod_log) / sigma(mod_log)
outliers_log <- design_df %>%
  filter(abs(std_resid_log) > 2.5) %>%
  select(participant, scales, polygon, log_ratio, std_resid_log)
```

### Signed Percent Error
```{r echo=FALSE}
pct_raw <- summary(mod_pct)[[1]]
pct_tbl <- data.frame(
  Df     = pct_raw$Df,
  SumSq  = round(pct_raw$`Sum Sq`, 3),
  MeanSq = round(pct_raw$`Mean Sq`, 3),
  Fvalue = round(pct_raw$`F value`, 3),
  Pvalue = round(pct_raw$`Pr(>F)`, 3),
  row.names = c("Scale", "Polygon", "Participant", "Residuals")
)
knitr::kable(pct_tbl,
             col.names = c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)"),
             caption = "ANOVA Table - Signed Percent Error",
             na_string = "-")
```
```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=4}
ggplot(data.frame(resid = residuals(mod_pct)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot - Signed Percent Error",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```
```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=4}
ggplot(data.frame(fitted = fitted(mod_pct),
                  resid  = residuals(mod_pct)),
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted - Signed Percent Error",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```
```{r echo=FALSE}
assump_pct <- data.frame(
  Test      = c("Shapiro-Wilk", "Levene (Scale)", "Levene (Polygon)"),
  Statistic = c(round(sw_pct$statistic, 4),
                round(lev_scale_pct[1,2], 4),
                round(lev_poly_pct[1,2], 4)),
  P.value   = c(round(sw_pct$p.value, 4),
                round(lev_scale_pct[1,3], 4),
                round(lev_poly_pct[1,3], 4)),
  Result    = c("Fail", "Pass", "Pass")
)
knitr::kable(assump_pct, caption = "Assumption Tests - Signed Percent Error")
```
```{r echo=FALSE}
knitr::kable(outliers, caption = "Detected Outliers (|Standardized Residual| > 2.5)")
```

Through our exploratory data analysis, and now formally, by examining standardized residual values, we identified a major outlier in the data. Participant 7 grossly overestimated the area of the concave polygon when shown at a medium scale, with an estimate over 600\% larger than the true value. With a standardized residual over 3, we decided to repeat analysis by omitting estimates from this participant. This decision was further supported by the mismatch we observed between the Q-Q plot and Shapiro-Wilk test of normality. The Q-Q plot indicated a pretty strong fit to a normal distribution, with the exclusion of one true point, and the Shapiro-Wilk test strongly indicated non-normality, invalidating our ANOVA results. If this outlying point had enough power to skew the Shapiro-Wilk test, its exclusion may reveal important insights in a revised ANOVA.

### Signed Percent Error (Without Outlier)
```{r echo=FALSE}
noP7_raw <- summary(mod_noP7)[[1]]
noP7_tbl <- data.frame(
  Df     = noP7_raw$Df,
  SumSq  = round(noP7_raw$`Sum Sq`, 3),
  MeanSq = round(noP7_raw$`Mean Sq`, 3),
  Fvalue = round(noP7_raw$`F value`, 3),
  Pvalue = round(noP7_raw$`Pr(>F)`, 3),
  row.names = c("Scale", "Polygon", "Participant", "Residuals")
)
knitr::kable(noP7_tbl,
             col.names = c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)"),
             caption = "ANOVA Table - Signed Percent Error (Without Outlier)",
             na_string = "-")
```
```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=4}
ggplot(data.frame(resid = residuals(mod_noP7)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot - Signed Percent Error (Without Outlier)",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```
```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=4}
ggplot(data.frame(fitted = fitted(mod_noP7),
                  resid  = residuals(mod_noP7)),
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted - Signed Percent Error (Without Outlier)",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```
```{r echo=FALSE}
assump_noP7 <- data.frame(
  Test      = c("Shapiro-Wilk", "Levene (Scale)", "Levene (Polygon)"),
  Statistic = c(round(sw_noP7$statistic, 4),
                round(lev_scale_noP7[1,2], 4),
                round(lev_poly_noP7[1,2], 4)),
  P.value   = c(round(sw_noP7$p.value, 4),
                round(lev_scale_noP7[1,3], 4),
                round(lev_poly_noP7[1,3], 4)),
  Result    = c("Pass", "Pass", "Pass")
)
knitr::kable(assump_noP7,
             caption = "Assumption Tests - Signed Percent Error (Without Outlier)")
```

The adjusted Shapiro-Wilk test confirms that with the exception of one outlying observation, the signed percentage error estimates were normally distributed and all model assumptions were satisfied. Shapiro-Wilk showed normality of residuals (W = 0.962, p = 0.168), and Levene's test showed constant variance across both scale (F = 0.516, p = 0.601) and polygon groups (F = 0.040, p = 0.961). However, the interpretation of the final ANOVA results did not change. Scale was still non-significant and polygon type was non-significant. The participant blocking factor remained highly significant (F = 5.979, p < 0.001), again confirming that individual differences in spatial reasoning ability dominate variation in estimation accuracy.

However, the need to omit valid observational data to perform a statistically valid analysis indicated that perhaps the signed percentage error was not the best metric for interpretation. Upon re-evaluation and research into literature, the use of a signed percentage error was revealed to be problematic. The most critical issue is the way the metric is bounded. Since the theoretical lowest value for a signed percentage error is -100\%, where the participant makes an estimate of 0 square miles, the theoretical highest value was infinite. In practice, this meant our data was incredibly asymmetrical.

Furthermore, from the branch of psychophysics, there is an enormous wealth of peer-reviewed science that states that humans do not interpret area linearly, and as such, it would be erroneous to analyze area estimations with just a signed percentage error. Stevens's Power Law, first proposed back in 1957, states that the perceived magnitude of a stimulus is proportional to the magnitude of the stimuli raised to some experiment-derived constant $n$:

$$\psi = k \cdot \phi^n$$

While testing area estimates of a projected square at varying magnitudes, his work yielded an $n$ of 0.7, which means that human perception of area is inherently non-linear (this has been supported by a number of preceding papers). By introducing the log ratio $\log(\hat{a}/a)$, we can derive a metric that has been empirically shown to best represent this phenomenon. We re-approached our analysis with this revised metric.

### Log Ratio
```{r echo=FALSE}
log_raw <- summary(mod_log)[[1]]
log_tbl <- data.frame(
  Df     = log_raw$Df,
  SumSq  = round(log_raw$`Sum Sq`, 3),
  MeanSq = round(log_raw$`Mean Sq`, 3),
  Fvalue = round(log_raw$`F value`, 3),
  Pvalue = round(log_raw$`Pr(>F)`, 3),
  row.names = c("Scale", "Polygon", "Participant", "Residuals")
)
knitr::kable(log_tbl,
             col.names = c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)"),
             caption = "ANOVA Table - Log Ratio Model",
             na_string = "-")
```
```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=4}
ggplot(data.frame(resid = residuals(mod_log)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "QQ Plot - Log Ratio Model",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```
```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=4}
ggplot(data.frame(fitted = fitted(mod_log),
                  resid  = residuals(mod_log)),
       aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted - Log Ratio Model",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```
```{r echo=FALSE}
assump_log <- data.frame(
  Test      = c("Shapiro-Wilk", "Levene (Scale)", "Levene (Polygon)"),
  Statistic = c(round(sw_log$statistic, 4),
                round(lev_scale_log[1,2], 4),
                round(lev_poly_log[1,2], 4)),
  P.value   = c(round(sw_log$p.value, 4),
                round(lev_scale_log[1,3], 4),
                round(lev_poly_log[1,3], 4)),
  Result    = c("Pass", "Pass", "Pass")
)
knitr::kable(assump_log, caption = "Assumption Tests - Log Ratio Model")
```
```{r echo=FALSE}
if(nrow(outliers_log) == 0){
  cat("No outliers detected under log ratio transformation.")
} else {
  knitr::kable(outliers_log, caption = "Outliers - Log Ratio Model")
}
```

Under the log ratio transformation of estimated areas, all model assumptions were satisfied without excluding data. Shapiro-Wilk showed normality of residuals (W = 0.975, p = 0.422), Levene's test indicated constant variance across scale (F = 0.182, p = 0.835) and polygon groups (F = 0.816, p = 0.449), and no observations passed the 2.5 standardized residual threshold for outlier detection. Scale was non-significant (F = 1.089, p = 0.351) and polygon type was non-significant (F = 0.215, p = 0.808). The participant blocking factor was highly significant (F = 5.518, p < 0.001), confirming that individual differences in spatial reasoning ability account for the dominant share of variation in estimation accuracy.

As a final check of data quality, we wanted to verify that the order in which participants viewed the maps did not affect their estimates. We had controlled for this in our aforementioned method for assigning treatments, but wanted additional validation. If our assumption of no learning effect was invalidated, we wanted to investigate what have occurred. This was performed by adding in the trial order as a covariate in the model, and then examining if the fitted coefficient was statistically significant.
```{r echo=FALSE}
mod_order <- aov(log_ratio ~ scales + polygon + participant + as.numeric(col),
                 data = design_df)

order_raw <- summary(mod_order)[[1]]
order_tbl <- data.frame(
  Df     = order_raw$Df,
  SumSq  = round(order_raw$`Sum Sq`, 3),
  MeanSq = round(order_raw$`Mean Sq`, 3),
  Fvalue = round(order_raw$`F value`, 3),
  Pvalue = round(order_raw$`Pr(>F)`, 3),
  row.names = c("Scale", "Polygon", "Participant", "Trial Order", "Residuals")
)
knitr::kable(order_tbl,
             col.names = c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)"),
             caption = "Model with Trial Order Covariate - Log Ratio",
             na_string = "-")
```

The model fit showed that trial order did not have a statistically significant effect on estimation accuracy, providing no evidence of systematic learning or fatigue across trials.


# Conclusions

This experiment investigated whether the scale at which a polygon is displayed on a map affects a participant's ability to accurately estimate its area. [ ]

## Limitations and Improvements
There were several limitations that should be taken into account. First, many participants were unfamiliar with square miles as a unit, likely introducing systematic overestimation bias unrelated to scale. Future studies could provide unit conversion references or use more intuitive units. Second, the 30-second time limit may have prevented deliberate estimation, capturing first impressions rather than considered judgments. Third, the shared library setting created potential for participants to overhear each other's guesses, which may have anchored some responses. Finally, with only 15 participants, the study was underpowered to detect small effects of scale. A larger sample with greater replication would improve statistical power considerably.

# References

Instituto Geográfico Militar de la República Argentina. (2009). 
POSGAR 2007 / Argentina 1 (EPSG:5343). EPSG Geodetic Parameter Registry. 
Retrieved from https://epsg.io/5343

# Appendices

-   `R` code.
