---
title: 'Final Project'
subtitle: "PSTAT122: Design and Analysis of Experiments" 

author:
  - name: "Winter 2026"
   # affiliations:
    #  - name: "Fall 2025"
affiliation-title: "Quarter"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    geometry: margin=0.2in

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)

```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Zachary Scholnick    (3374824)
- STUDENT 2    (NetID 2)
- STUDENT 3    (NetID 3)
- STUDENT 4    (NetID 4)
:::


::: {.callout-caution}
# Due Date
**Due Date:** Monday, March 16, 2026, 11:59 PM
:::


# Introduction
-	Clear statement of the objective or research question.
-	Brief context or motivation.

In the field of map-making and cartography, scale is one of the most critical and challenging decisions practitioners face. At too large of a scale (zoomed in), key features are easily visible, but lose out on important context. At the same time, a very low scale (zoomed out), means that the information is difficult to see and fine spatial details are lost. Getting the scale of your map right is critical in conveying key information - and for that reason it's a question asked every time a map is created. On top of the practical aspects of mapmaking, humans possess a number of cognitive biases that make interpretations of proportion and area very difficult. As such, this experiment was motivated by the desire to provide real insights into what scale of representation makes estimation of area most intuitive.

While tackling this in entirety is beyond the scope of this project, we chose to focus on testing a key component of this issue: How does the scale of a mapped polygon affect the accuracy in which users are able to estimate the area? Do people systematically overestimate or underestimate polygon area at different scales?


# Experimental Design
- Description of factors and treatment structure.
- Clearly state what you are measuring and the units.
Examples: Number of words recalled (count), reaction time (seconds), taste rating (1–5 scale).
- Identify which factors are fixed vs. random.
- Description of design type (CRD, RCBD, factorial, etc.).
- Explain how randomization, replication, and (if used) blocking were implemented.
- Sample size: Provide number of observations per condition.
Guideline: 5–10 per treatment for CRD, 3–5 blocks for RCBD, total feasible within 1 hour.

(You are encouraged to explore more resources for determining the sample size )

## Experimental Design

To assess this, we created 3 polygons of equal area, 1.71 square miles, selected because mapped parcels, agricultural units, and other targets of interest are commonly in the 1-3 square mile range. These 3 polygons represented common archetypes in geometric representation: the simple polygon, rounded polygons, and the more complex, concave polygon. We selected 3 random regions in the Patagonian Steppe in Argentina, as this region lacked easily recognizable reference points of scale, like trees or streams, and consisted mainly of rock formations of unidentifiable size. To ensure consistent areal calculation of mapped region and polygon, we opted for the POSGAR 2007/EPSG:5343 projection, which is the official geodetic reference system of Argentina.

Each polygon was then plotted at the scale ratios of 1:13,390, 1:26,780, and 1:59,852 to provide a large-scale, medium-scale, and small-scale plot. In a more generalizable interpretation, this means that we plotted polygons such that they covered 40%, 10%, and 2% of the entire mapped region. The scale bar itself was dynamically recalculated using QGIS based on the current map extent with a fixed display size for a constant visual across maps.

| Scale Label | Scale Ratio | Polygon Coverage |
|-------------|-------------|-----------------|
| Large (L)   | 1:13,390    | 40%             |
| Medium (M)  | 1:26,780    | 10%             |
| Small (S)   | 1:59,852    | 2%              |

: Scale treatments and polygon coverage of map frame {#tbl-scales}

To isolate the desired effect of the impact of scale on area estimates, we employed the latin squares experimental design to account for variations within user interpretation and type of polygon. We expected some participants to systematically perform at higher or lower levels of success across every presented task dependent on a number of external factors, including eyesight and familiarity with related tasks. By controlling for participants, we seek to isolate common patterns across the studied population. We also hypothesized that the different shapes of polygons would modify the difficulty of area estimates at varying scales. By introducing 3 archetypal polygons, and recording results per polygon, we can also account for the variation in performance induced by polygon type. This brings us to the following standard latin square setup:

| | Curved Polygon | Simple Polygon | Concave Polygon |
|-------------|----------------|----------------|-----------------|
| **Participant 1** | S | M | L |
| **Participant 2** | M | L | S |
| **Participant 3** | L | S | M |

: Latin square treatment assignment for one replicate block {#tbl-latin}

In this experiment, each participant was given 30 seconds to write down an area estimate in square miles of a curved, simple and concave polygon, each shown at a unique scale. This response was then transformed into a signed percent error:

$$\text{Percent Error} = \frac{\hat{a} - a}{a} \times 100$$

where $\hat{a}$ is the participant's estimate and $a = 1.71$ square miles is the true polygon area. While each polygon has equal size, the participants were not informed during the experiment. 5 replications were performed, and with 3 participants per block, the total experiment had 15 participants. All factors in the experiment - polygon type, participant, and scale - were fixed effects.

# Data Collection
- **Procedure**: Describe how and when the experiment was conducted (e.g., location, date, steps taken).

Slides for running trials: https://docs.google.com/presentation/d/1cQkHKFSfIFLclCvy6DIQ8lRVRss23OmEZ6OrBmpAJ2I/edit?usp=sharing

```{r}
library(agricolae)

scales <- c("L", "M", "S")
polygon_types <- c("curved", "simple", "concave")

# 1. Map each participant to a specific replicate AND a specific row
set.seed(42)
participant_assignment <- data.frame(
  replicate = rep(1:5, each = 3),
  row = rep(1:3, times = 5),
  # Randomly shuffle participants P1 through P15 into these slots
  participant = paste0("P", sample(1:15)) 
)

# 2. Generate the 5 Latin Squares
design_df <- do.call(rbind, lapply(1:5, function(i) {
  book <- design.lsd(scales, seed = i * 10)$book
  book$replicate <- i
  
  # Assign polygon types to columns (1, 2, 3)
  set.seed(i)
  col_map <- setNames(sample(polygon_types), 1:3)
  book$polygon <- col_map[as.character(book$col)]
  
  return(book)
}))

# 3. THE FIX: Merge on BOTH replicate and row 
design_df <- merge(design_df, participant_assignment, by = c("replicate", "row"))

# 4. Sort neatly by participant (P1 to P15) and then by trial/column (1, 2, 3)
# (using a custom sort so P10 doesn't jump in front of P2)
design_df <- design_df[order(as.numeric(gsub("P", "", design_df$participant)), design_df$col), 
                       c("participant", "replicate", "row", "col", "scales", "polygon")]

# Add empty response column
design_df$percent_error <- NA

cat("\nFull Design\n")
print(design_df)
```


```{r}
enter_data <- function(df, p_name, estimates) {
  # Quick check to make sure you entered exactly 3 values
  if(length(estimates) != 3) {
    stop("Please provide exactly 3 estimates for the participant.")
  }
  
  # Calculate the percent error based on the true area (1.71 sq miles)
  errors <- ((estimates - 1.71) / 1.71) * 100
  
  # Assign the errors to col 1, 2, and 3 (Trial 1, 2, and 3) for that specific participant
  for(i in 1:3) {
    df$percent_error[df$participant == p_name & df$col == i] <- errors[i]
  }
  
  return(df)
}

# --- Example Usage ---
# Enter data for P1, P2, and P3 as you collect it:
design_df <- enter_data(design_df, "P1", c(9.8, 6.5, 12.0))
design_df <- enter_data(design_df, "P2", c(3, 3, 3))
design_df <- enter_data(design_df, "P3", c(2, 1, 3.2))
design_df <- enter_data(design_df, "P4", c(3, 4, 4.5))
design_df <- enter_data(design_df, "P5", c(3.1, 2.8, 2.6))
design_df <- enter_data(design_df, "P6", c(2.4, 2.2, 1.6))
design_df <- enter_data(design_df, "P7", c(5, 12, 6))

```

- **Challenges/Adjustments**: Mention any difficulties or changes made during data collection (e.g., technical issues, time adjustments).
- **Data Presentation**: Display the collected data in tables or graphs, summarizing key measures like mean and standard deviation..

# Analysis

- **Exploratory Data**: Start with basic statistics (mean, SD) and visualizations (e.g., boxplots) to understand the data.


- **Hypothesis Testing**: Test your hypothesis with an appropriate statistical test (e.g., ANOVA).

- **Tables, Figures, & Code**: Include key results (ANOVA table, post-test) and relevant `R` code excerpts where needed.

-	Use `R` to analyze the data.

# Conclusions
-	Summarize key findings.
- Comment on limitations and possible improvements.


# References
 (If needed.)

# Appendices
 
 - `R` code.


