---
title: 'Final Project'
subtitle: "PSTAT122: Design and Analysis of Experiments" 

author:
  - name: "Winter 2026"
   # affiliations:
    #  - name: "Fall 2025"
affiliation-title: "Quarter"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    geometry: margin=0.2in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)

```

:::: callout
::: {style="text-align: center"}
[ **STUDENT NAME** ]{style="color: blue;"}
:::

-   Zachary Scholnick (3374824)
-   Anjali Dev (A167A78)
-   Logan Marks (9536947)

::::

::: callout-caution
# Due Date

**Due Date:** Monday, March 16, 2026, 11:59 PM
:::

# Introduction

In the field of map-making and cartography, scale is one of the most critical and challenging decisions practitioners face. At too large of a scale (zoomed in), key features are easily visible, but lose out on important context. At the same time, a very low scale (zoomed out), means that the information is difficult to see and fine spatial details are lost. Getting the scale of your map right is critical in conveying key information - and for that reason it's a question asked every time a map is created. On top of the practical aspects of mapmaking, humans possess a number of cognitive biases that make interpretations of proportion and area very difficult. As such, this experiment was motivated by the desire to provide real insights into what scale of representation makes estimation of area most intuitive.

While tackling this in entirety is beyond the scope of this project, we chose to focus on testing a key component of this issue: How does the scale of a mapped polygon affect the accuracy in which users are able to estimate the area? Do people systematically overestimate or underestimate polygon area at different scales?


## Experimental Design

To assess this, we created 3 polygons of equal area, 1.71 square miles, selected because mapped parcels, agricultural units, and other targets of interest are commonly in the 1-3 square mile range. These 3 polygons represented common archetypes in geometric representation: the simple polygon, rounded polygons, and the more complex, concave polygon. We selected 3 random regions in the Patagonian Steppe in Argentina, as this region lacked easily recognizable reference points of scale, like trees or streams, and consisted mainly of rock formations of unidentifiable size. To ensure consistent areal calculation of mapped region and polygon, we opted for the POSGAR 2007/EPSG:5343 projection, which is the official geodetic reference system of Argentina.

Each polygon was then plotted at the scale ratios of 1:13,390, 1:26,780, and 1:59,852 to provide a large-scale, medium-scale, and small-scale plot. In a more generalizable interpretation, this means that we plotted polygons such that they covered 40%, 10%, and 2% of the entire mapped region. The scale bar itself was dynamically recalculated using QGIS based on the current map extent with a fixed display size for a constant visual across maps.

| Scale Label | Scale Ratio | Polygon Coverage |
|-------------|-------------|------------------|
| Large (L)   | 1:13,390    | 40%              |
| Medium (M)  | 1:26,780    | 10%              |
| Small (S)   | 1:59,852    | 2%               |

: Scale treatments and polygon coverage of map frame {#tbl-scales}

To isolate the desired effect of the impact of scale on area estimates, we employed the latin squares experimental design to account for variations within user interpretation and type of polygon. We expected some participants to systematically perform at higher or lower levels of success across every presented task dependent on a number of external factors, including eyesight and familiarity with related tasks. By controlling for participants, we seek to isolate common patterns across the studied population. We also hypothesized that the different shapes of polygons would modify the difficulty of area estimates at varying scales. By introducing 3 archetypal polygons, and recording results per polygon, we can also account for the variation in performance induced by polygon type. This brings us to the following standard latin square setup:

|                   | Curved Polygon | Simple Polygon | Concave Polygon |
|-------------------|----------------|----------------|-----------------|
| **Participant 1** | S              | M              | L               |
| **Participant 2** | M              | L              | S               |
| **Participant 3** | L              | S              | M               |

: Latin square treatment assignment for one replicate block {#tbl-latin}

Scale treatments were randomly assigned to rows and columns within each Latin square replicate using R's agricolae package with a fixed seed for reproducibility. In this experiment, each participant was given 30 seconds to write down an area estimate in square miles of a curved, simple and concave polygon, each shown at a unique scale. This response was then transformed into a signed percent error:

$$\text{Percent Error} = \frac{\hat{a} - a}{a} \times 100$$

where $\hat{a}$ is the participant's estimate and $a = 1.71$ square miles is the true polygon area. While each polygon has equal size, the participants were not informed during the experiment. 5 replications were performed, and with 3 participants per block, the total experiment had 15 participants. With 5 replicates of 3 participants each, this yields 5 observations per scale treatment, falling within the recommended guideline of 5–10 per treatment. All factors: Scale, polygon type, and participant block were all treated as fixed effects, as each level was deliberately selected rather than sampled from a larger population.


```{r}
library(agricolae)

scales <- c("L", "M", "S")
polygon_types <- c("curved", "simple", "concave")

# 1. Map each participant to a specific replicate AND a specific row
set.seed(42)
participant_assignment <- data.frame(
  replicate = rep(1:5, each = 3),
  row = rep(1:3, times = 5),
  # Randomly shuffle participants P1 through P15 into these slots
  participant = paste0("P", sample(1:15)) 
)

# 2. Generate the 5 Latin Squares
design_df <- do.call(rbind, lapply(1:5, function(i) {
  book <- design.lsd(scales, seed = i * 10)$book
  book$replicate <- i
  
  # Assign polygon types to columns (1, 2, 3)
  set.seed(i)
  col_map <- setNames(sample(polygon_types), 1:3)
  book$polygon <- col_map[as.character(book$col)]
  
  return(book)
}))

# 3. THE FIX: Merge on BOTH replicate and row 
design_df <- merge(design_df, participant_assignment, by = c("replicate", "row"))

# 4. Sort neatly by participant (P1 to P15) and then by trial/column (1, 2, 3)
# (using a custom sort so P10 doesn't jump in front of P2)
design_df <- design_df[order(as.numeric(gsub("P", "", design_df$participant)), design_df$col), 
                       c("participant", "replicate", "row", "col", "scales", "polygon")]

# Add empty response column
design_df$percent_error <- NA

cat("\nFull Design\n")
print(design_df)
```

# Data Collection

**Procedure**: Each participant was given 30 seconds to estimate the area in square miles of each polygon shown. Trials were conducted using a slide presentation displayed on a laptop screen, with each slide showing one polygon at its assigned scale alongside a scale bar. Participants recorded their estimates on paper. All 5 replicates were conducted in the UCSB library on March 14, 2026. Participants were seated individually and shown the slides in sequence, with no prior knowledge of the true polygon area or that all three polygons were equal in size.

Slides for running trials: https://docs.google.com/presentation/d/1cQkHKFSfIFLclCvy6DIQ8lRVRss23OmEZ6OrBmpAJ2I/edit?usp=sharing

**Challenges/Adjustments**: Several challenges arose during data collection. The most notable was that some participants were unfamiliar with square miles as a unit of measurement, which may have introduced systematic bias into estimates regardless of scale treatment. While participants were not coached on how to use the scale bar, some may have relied on intuition rather than calculation. Additionally, the 30-second time limit, while intentional in capturing instinctive responses, may have prevented participants from making more deliberate estimates — meaning results may reflect first impressions of area rather than considered judgments. Finally, as trials were conducted in a shared library space, there is a small possibility that some participants overheard the estimates of others, which could have anchored subsequent guesses. No adjustments were made to the protocol during data collection.

**Data Presentation**: 

```{r}
enter_data <- function(df, p_name, estimates) {
  # Quick check to make sure you entered exactly 3 values
  if(length(estimates) != 3) {
    stop("Please provide exactly 3 estimates for the participant.")
  }
  
  # Calculate the percent error based on the true area (1.71 sq miles)
  errors <- ((estimates - 1.71) / 1.71) * 100
  
  # Assign the errors to col 1, 2, and 3 (Trial 1, 2, and 3) for that specific participant
  for(i in 1:3) {
    df$percent_error[df$participant == p_name & df$col == i] <- errors[i]
  }
  
  return(df)
}

#data collection 
design_df <- enter_data(design_df, "P1", c(4.8, 4.5, 3.0))
design_df <- enter_data(design_df, "P2", c(3, 3, 3))
design_df <- enter_data(design_df, "P3", c(2, 1, 3.2))
design_df <- enter_data(design_df, "P4", c(3, 4, 4.5))
design_df <- enter_data(design_df, "P5", c(3.1, 2.8, 2.6))
design_df <- enter_data(design_df, "P6", c(2.4, 2.2, 1.6))
design_df <- enter_data(design_df, "P7", c(5, 12, 6))
design_df <- enter_data(design_df, "P8", c(2.4, 2.5, 2))
design_df <- enter_data(design_df, "P9", c(1.43, 2.8, 3.2))
design_df <- enter_data(design_df, "P10", c(3.2, 3.5, 4.1))
design_df <- enter_data(design_df, "P11", c(2.7, 2.9, 3.1))
design_df <- enter_data(design_df, "P12", c(4.3, 4.1, 3.9))
design_df <- enter_data(design_df, "P13", c(3.6, 3.2, 3.8))
design_df <- enter_data(design_df, "P14", c(2.9, 2.4, 2.7))
design_df <- enter_data(design_df, "P15", c(4.8, 4.5, 4.9))

knitr::kable(design_df, 
             caption = "Full experimental data with percent error by participant, polygon type, and scale",
             digits = 2)
```


```{r}
library(dplyr)

summary_table <- design_df %>%
  group_by(scales) %>%
  summarise(
    Mean_Error = round(mean(percent_error, na.rm = TRUE), 2),
    SD_Error = round(sd(percent_error, na.rm = TRUE), 2),
    N = n()
  )

knitr::kable(summary_table,
             
             col.names = c("Scale", "Mean % Error", "SD % Error", "N"))
```


# Analysis

**Exploratory Data**:
```{r}
library(dplyr)
library(ggplot2)

# Basic summary statistics by scale
summary_stats <- design_df %>%
  group_by(scales) %>%
  summarise(
    Mean = round(mean(percent_error, na.rm = TRUE), 2),
    SD = round(sd(percent_error, na.rm = TRUE), 2),
    Min = round(min(percent_error, na.rm = TRUE), 2),
    Max = round(max(percent_error, na.rm = TRUE), 2),
    N = n()
  )

knitr::kable(summary_stats, caption = "Summary statistics of percent error by scale treatment")

# Boxplot by scale
ggplot(design_df, aes(x = scales, y = percent_error, fill = scales)) +
  geom_boxplot() +
  labs(title = "Percent Error by Scale Treatment",
       x = "Scale", y = "Percent Error (%)") +
  theme_minimal()

# Boxplot by polygon type
ggplot(design_df, aes(x = polygon, y = percent_error, fill = polygon)) +
  geom_boxplot() +
  labs(title = "Percent Error by Polygon Type",
       x = "Polygon Type", y = "Percent Error (%)") +
  theme_minimal()
```

The boxplots show that percent error is fairly similar across all three scale treatments (Large, Medium, Small), with most participants overestimating the polygon area by around 60–100%. There is one clear outlier with a percent error near 600%, which heavily influences the spread. Across polygon types, concave polygons had the most consistent estimates while curved polygons showed the most variation. Overall, the differences between groups appear modest, which we will test formally using ANOVA.



**Hypothesis Testing**: Test your hypothesis with an appropriate statistical test (e.g., ANOVA).
```{r}
# Latin square ANOVA model
model <- aov(percent_error ~ scales + polygon + participant, data = design_df)
summary(model)

TukeyHSD(model, "scales")
```

We test the following hypotheses:
$$H_0: \mu_L = \mu_M = \mu_S$$

$$ H_a: \text{At least one scale mean differs} $$
where $$\mu_L$$, $$\mu_M$$, and $$\mu_S$$ represent the mean percent error under large, medium, and small scale treatments respectively. A Latin square ANOVA was fitted with scale as the primary treatment factor and polygon type and participant as blocking factors.

The ANOVA results show that scale treatment was not statistically significant (F = 0.144, p = 0.867), meaning we fail to reject the null hypothesis. There is no evidence that map scale affects the accuracy of area estimation. Similarly, polygon type was not significant (F = 0.258, p = 0.774), suggesting that polygon shape did not meaningfully influence estimation accuracy.

However, the participant blocking factor was highly significant (F = 4.303, p < 0.001), confirming that individual participants varied substantially in their estimation ability. This validates our decision to use a Latin square design to control for participant-level variation — without blocking, this variability would have inflated our residual error and made it even harder to detect a scale effect.

The Tukey HSD post-hoc test confirmed no significant pairwise differences between any scale pairs (M-L: p = 0.862, S-L: p = 0.927, S-M: p = 0.988), with all confidence intervals comfortably overlapping zero.

**Tables, Figures, & Code**:

```{r}
# ANOVA Table
model <- aov(percent_error ~ scales + polygon + participant, data = design_df)
knitr::kable(summary(model)[[1]], 
             caption = "Latin Square ANOVA Results",
             digits = 3)

# Tukey Post-hoc Table
tukey_results <- TukeyHSD(model, "scales")
tukey_df <- as.data.frame(tukey_results$scales)
knitr::kable(tukey_df,
             caption = "Tukey HSD Post-hoc Test for Scale Treatment",
             digits = 3,
             col.names = c("Difference", "Lower 95% CI", "Upper 95% CI", "Adjusted p-value"))
```


# Conclusions

This experiment investigated whether the scale at which a polygon is displayed on a map affects a participant's ability to accurately estimate its area. Based on our Latin square ANOVA, we found no statistically significant effect of map scale on percent error (F = 0.144, p = 0.867), nor of polygon type (F = 0.258, p = 0.774). As such, we fail to reject the null hypothesis — there is no evidence that large, medium, or small scale presentations lead to meaningfully different estimation accuracy. Notably, all participants systematically overestimated polygon area regardless of scale or shape, suggesting a general human tendency to overestimate mapped area.

The most significant finding was the strong participant effect (F = 4.303, p < 0.001), confirming that individual differences in spatial reasoning ability account for far more variation in estimation accuracy than the experimental treatments themselves. This validates the Latin square blocking structure used in the design.

## Limitations and Improvements
Several limitations should be noted. First, many participants were unfamiliar with square miles as a unit, likely introducing systematic overestimation bias unrelated to scale. Future studies could provide unit conversion references or use more intuitive units. Second, the 30-second time limit may have prevented deliberate estimation, capturing first impressions rather than considered judgments. Third, the shared library setting created potential for participants to overhear each other's guesses, which may have anchored some responses. Finally, with only 15 participants, the study was underpowered to detect small effects of scale. A larger sample with greater replication would improve statistical power considerably.

# References

Instituto Geográfico Militar de la República Argentina. (2009). 
POSGAR 2007 / Argentina 1 (EPSG:5343). EPSG Geodetic Parameter Registry. 
Retrieved from https://epsg.io/5343

# Appendices

-   `R` code.
